---
title: "Bridger-Teton National Forest RAP Data Summary"
author: "Matthew Van Scoyoc"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
format: 
  docx:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    reference-doc: "style.docx"
editor_options: 
  chunk_output_type: inline
execute:
  echo: false
  message: false
  warning: false
  class-output: 'r'
  fig-width: 6.5
---

```{r proj_vars}
# Project Variables ----
# NOTE: change the variables in this section and run the rest of the script.

# Forest name
unit_name <- "Bridger-Teton NF"

# Path to project directory
proj_dir <- file.path("T:/FS/NFS/PSO/MPSG/Workspace/MVanScoyoc_GISLead/BT_RAP")
# Figures directory path 
fig_dir <- file.path(proj_dir, "figures")
# Raster directory path
raster_dir <- file.path(proj_dir, "rap_data")
# Years of RAP data
rap_yrs <- c(1986:2024)

#-- Spatial variable
# Project coordinate reference system (crs)
target_crs <- "EPSG:4326" # crs of raster data# Path to geodatabase
# Geodatabase path
gdb_path <- file.path(proj_dir, "BT_RAP.gdb")
# Area of Interest feature class names
aoi_fc <- "btnf_admin_bndry"
ecos_full_fc <- "btnf_nonforest_ecosystems"
ecos_diss_fc <- "btnf_nf_ecos_dissolve"
```

```{r load_data, eval=TRUE}
load(file.path(proj_dir, "RAP_summary_data.RData"))
```

```{r setup}
## Packages ----
#-- list of packages
pkgs <- c(
  "cowplot",     # Plot tools
  "ggspatial",   # Map tools
  "ggtext",      # Plot tools
  "prettymapr",  # Map tools
  "remotes",     # Download packages from GitHub
  "scales",      # Map tools
  "sf",          # Spatial vector data tools
  "stringr",     # Text string tools 
  "terra",       # Raster data tools
  "tidyterra",   # Map tools
  "tidyverse"    # Data management tools
)

#-- install packages
inst_pkgs <- pkgs %in% rownames(installed.packages())
# rapr package is on GitHub
if (!"rapr" %in% inst_pkgs) remotes::install_github("brownag/rapr")
# everything else is on CRAN
if (any(inst_pkgs == FALSE)) {
  install.packages(pkgs[!inst_pkgs], 
                   lib =  .libPaths()[1], 
                   repos = "https://cloud.r-project.org",
                   type = 'source', 
                   dependencies = TRUE, 
                   quiet = TRUE)
}

#-- load packages
invisible(lapply(pkgs, library, character.only = TRUE))


# Functions ----

#' Create change maps
#' This function is used in an `lapply()` function to produce a list of change 
#'     maps.
#'
#' @param lyr Name of raster layer. 
#' @param sub_rast Subsetted raster
#'
#' @return A ggplot object.
change_fig <- function(lyr, sub_rast, legend_title){
  # Calculate 5 and 95th percentile
  sub_rast_df <- as.data.frame(sub_rast)
  lowbound <- quantile(sub_rast_df$mean, probs = 0.05, na.rm = T)[[1]]
  upbound <- quantile(sub_rast_df$mean, probs = 0.95, na.rm = T)[[1]]
  # Plot data
  ggplot() +
    tidyterra::geom_spatraster(data = sub_rast) +
    scale_fill_gradient2(
      low = "#DC050C", mid = "#FFFFFF", high = "#1965B0", 
      na.value = "transparent",
      midpoint = 0, limits = c(lowbound, upbound), oob = scales::squish,
      name = expression(legend_title)
    ) +
    tidyterra::geom_spatvector(data = aoi_bndry, fill = "transparent", 
                               color = "black") +
    theme_minimal() +
    ggspatial::annotation_scale(location = "br", width_hint = 0.5,
                                pad_x = unit(0.1, "cm"), 
                                pad_y = unit(0.1, "cm")) +
    ggspatial::annotation_north_arrow(
      location = "tr", style = ggspatial::north_arrow_minimal(),
      height = unit(0.7, "cm"), width = unit(0.7, "cm"),
      pad_x = unit(0.1, "cm"), pad_y = unit(0.1, "cm")
    ) +
    theme(legend.justification.right = "top", axis.text = element_blank(),
          axis.ticks = element_blank(), axis.title = element_blank()) +
    ggtitle(paste0(stringr::str_to_title(lyr)))
}


#' Concatinate a text string for use in a paragraph.
#'
#' This funciton places commans and "and" in a text sting vector so it reads 
#'     well.
#'
#' @param text_vector A vector of text strings.
#'
#' @return A vector.
commas_in_string <- function(text_vector){
  # text_vector = ecosystems_df$BPS_NAME
   if (length(text_vector) == 1){
     text_vector
     } else if (length(text_vector) > 2){
       paste0(
         str_c(text_vector[1:length(text_vector)-1], collapse = ", "),
         ", and ",
         text_vector[length(text_vector)]
         )
       } else if (length(text_vector == 2)){
         paste0(text_vector[1], " and ", text_vector[2])
       }
  }
```


# Introduction

The [Rangeland Analysis Platform](https://rangelands.app/) (RAP) hosts above-ground biomass and plant cover estimates derived from LANDSAT imagery.
RAP above-ground biomass and plant cover data (USDA Agricultural Research Service 2025) from `r min(rap_yrs)` to `r max(rap_yrs)` were downloaded on 2025-05-20 using the `rpar` package (version 1.0.0; Brown 2025) in R (version `r paste(R.version$major, R.version$minor, sep = ".")`; R Core Team `r R.version$year`).


# Area of Analysis

```{r subset_vectors}
alpine <- ecosystems[ecosystems$BT_Type == "Alpine", ]
sage <- ecosystems[ecosystems$BT_Type == "Sagebrush", ]

# Convert ecosystem feature class to a data frame 

alpine_bps <- dplyr::filter(ecosystems_df, BT_Type == "Alpine") |> 
  dplyr::pull(BPS_NAME)
sage_bps <- dplyr::filter(ecosystems_df, BT_Type == "Sagebrush") |> 
  dplyr::pull(BPS_NAME)
```

Alpine and sagebrush ecosystems were identified by `r unit_name` staff as non-forested systems in need of additional data to help inform decision making.
LANDFIRE Bio-physical Setting (BpS) data were used to identify the spatial extent of these ecosystems on the Forest (Figure 1);
`r commas_in_string(alpine_bps)` were selected for alpine ecosystems, and
`r commas_in_string(sage_bps)` was selected for sagebrush ecosystems.
RAP data were downloaded for the entire administrative forest boundary and then clipped to the alpine and sagebrush ecosystems for these analyses.

```{r aoi_map}
#| fig-cap: "Figure 1. Map of the Bridger-Teton NF showing alpine and sagebrush ecosystems."
#| fig-alt: "Figure 1. Map of the Bridger-Teton NF showing alpine and sagebrush ecosystems."
#| fig-height: 6.5

# plot quick map
ggplot() +
  ggspatial::annotation_map_tile(type = "osm", zoom = 8, alpha = 0.3) +
  geom_sf(data = ecosystems, aes(fill = BT_Type), color = "transparent") +
  scale_fill_manual(values = c("#1965B0", "#DC050C")) +
  geom_sf(data = aoi_bndry, fill = "transparent", color = "black") +
  theme_minimal() +
  annotation_scale(location = "br", width_hint = 0.5,
                   pad_x = unit(0.1, "cm"), pad_y = unit(0.1, "cm")) +
  annotation_north_arrow(location = "tr", style = north_arrow_minimal(),
                         height = unit(0.7, "cm"), width = unit(0.7, "cm"),
                         pad_x = unit(0.1, "cm"), pad_y = unit(0.1, "cm")) +
  theme(legend.justification.right = "top") +
  labs(title = paste0("Area of Interest - ", unit_name), 
       fill = "Ecosystems") +
  xlab("Longitude") + ylab("Latitude")
```


{{< pagebreak >}}
# Biomass Data {#sec-biomass}

Below is a figure of average (mean) annual above-ground forb and grass biomass (lbs/acre) by lifespan (annual/perennial) from `r min(rap_yrs)` to `r max(rap_yrs)` for alpine and sagebrush ecosystems (Figure 2).

```{r bm_ts}
#| fig-cap: "Figure 2. Time-series of mean percent cover by life for alpine and sagebrush ecosystems."
#| fig-alt: "Figure 2. Time-series of mean annual forb and grass biomass by lifespan for alpine and sagebrush ecosystems."

## Plot Time-series ----
ggplot(mean_bm_ts, aes(x = year, y = mean)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 0.5) +  
  facet_grid(type ~ ecosystem, scales = "free_y") +
  labs(title = "Time-series of Mean Forb and Grass Biomass", 
       x = "Year", y = "Biomass (lbs/acre)") +
  theme_bw() +
  theme(strip.background = element_rect(fill = 'white'))
```

```{r bm_change, results='hide'}
#| output: false
#| include: false

#-- Alpine
bm_alp_chng <- terra::rast(
  file.path(
    raster_dir,
    paste0("RAP_biomass_change_alpine", min(rap_yrs), "_", max(rap_yrs), ".tif")
    )
  )
# Create figures
bm_alp_figs <- lapply(1:nlyr(bm_alp_chng), function(lyr){
  sub_rast = bm_alp_chng[[lyr]]
  my_var = stringr::str_split_i(names(sub_rast), "_", 2)
  if(my_var == "annual") my_var = "Annual Forb & Grass"
  if(my_var == "perennial") my_var = "Perennial Forb & Grass"
  print(my_var)
  p = change_fig(my_var, sub_rast, "Biomass (lbs/acre)")
  return(p)
})

#-- Sage
bm_sage_chng <- terra::rast(
  file.path(
    raster_dir,
    paste0("RAP_biomass_change_sage", min(rap_yrs), "_", max(rap_yrs), ".tif"))
)
# Create figures
bm_sage_figs <- lapply(1:nlyr(bm_sage_chng), function(lyr){
  sub_rast = bm_sage_chng[[lyr]]
  my_var = stringr::str_split_i(names(sub_rast), "_", 2)
  if(my_var == "annual") my_var = "Annual Forb & Grass"
  if(my_var == "perennial") my_var = "Perennial Forb & Grass"  
  print(my_var)
  p = change_fig(my_var, sub_rast, "Biomass (lbs/acre)")
  return(p)
})
```

```{r bm_alp_change}
#| fig-cap: "Figure 3. Change in above ground carbon in alpine ecosystems."
#| fig-alt: "Figure 3. Change in above ground carbon in alpine ecosystems."
#| fig-height: 8

# Final figure
gridExtra::grid.arrange(
  grobs = bm_alp_figs,
  ncol = 1,
  top = grid::textGrob("Change in Aboveground Carbon in Alpine Ecosystmes\n2020-2023 average minus 1990-1993 average",
                       gp = grid::gpar(fontsize = 14, fontface = "bold"))
)
```

```{r bm_sage_change}
#| fig-cap: "Figure 4. Change in above ground carbon in sagebrush ecosystems."
#| fig-alt: "Figure 4. Change in above ground carbon in sagebrush ecosystems."
#| fig-height: 8.5

# Final figure
gridExtra::grid.arrange(
  grobs = bm_sage_figs,
  nrow = 2,
  top = grid::textGrob("Change in Aboveground Carbon in Sagebrush Ecosystems\n2020-2023 average minus 1990-1993 average",
                       gp = grid::gpar(fontsize = 14, fontface = "bold"))
)
```


{{< pagebreak >}}
# Cover Data {#sec-cover}

Below is a figure of average (mean) percent vegetation cover by lifeform from `r min(rap_yrs)` to `r max(rap_yrs)` for alpine and sagebrush ecosystems (Figure 3).
```{r cov_fig}
#| fig-cap: "Figure 5. Time-series of mean percent cover by plant lifeform for alpine and sagebrush ecosystems."
#| fig-alt: "Figure 5. Time-series of mean percent cover by plant lifeform for alpine and sagebrush ecosystems."
#| fig-height: 6.5

## Plot Time-series ----
mean_cov_ts[mean_cov_ts$type == "Bare", "type"] <- "Bare Ground"
ggplot(mean_cov_ts, aes(x = year, y = mean)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 0.5) +  
  facet_grid(type ~ ecosystem, scales = "free_y") +
  labs(title = "Time-series of Mean Annual Percent Cover", 
       x = "Year", y = "Biomass (lbs/acre)") +
  theme_bw() +
  theme(strip.background = element_rect(fill = 'white'))
```

```{r cov_change, results='hide'}
#| output: false
#| include: false

#-- Alpine
cov_alp_chng <- terra::rast(
  file.path(
    raster_dir,
    paste0("RAP_cover_change_alpine", min(rap_yrs), "_", max(rap_yrs), ".tif")
    )
  )
# Create figures
cov_alp_figs <- lapply(1:nlyr(cov_alp_chng), function(lyr){
  sub_rast = cov_alp_chng[[lyr]]
  my_var = stringr::str_split_i(names(sub_rast), "_", 2)
  if(my_var == "annual") my_var = "Annual Forb & Grass"
  if(my_var == "perennial") my_var = "Perennial Forb & Grass"
  if(my_var == "bare") my_var = "Bare Ground"
  print(my_var)
  p = change_fig(my_var, sub_rast, "Percent Cover")
  return(p)
})

#-- Sage
cov_sage_chng <- terra::rast(
  file.path(
    raster_dir,
    paste0("RAP_cover_change_sage", min(rap_yrs), "_", max(rap_yrs), ".tif"))
)
# Create figures
cov_sage_figs <- lapply(1:nlyr(cov_sage_chng), function(lyr){
  sub_rast = cov_sage_chng[[lyr]]
  my_var = stringr::str_split_i(names(sub_rast), "_", 2)
  if(my_var == "annual") my_var = "Annual Forb & Grass"
  if(my_var == "perennial") my_var = "Perennial Forb & Grass"
  if(my_var == "bare") my_var = "Bare Ground"
  print(my_var)
  p = change_fig(my_var, sub_rast, "Percent Cover")
  return(p)
})
```

```{r cov_alp_change}
#| fig-cap: "Figure 6. Change in percent cover in alpine ecosystems."
#| fig-alt: "Figure 6. Change in percent cover in alpine ecosystems."
#| fig-height: 8.5

# Final figure
gridExtra::grid.arrange(
  grobs = cov_alp_figs,
  ncol = 2,
  top = grid::textGrob("Change in Percent Cover in Alpine Ecosystmes\n2020-2023 average minus 1990-1993 average",
                       gp = grid::gpar(fontsize = 14, fontface = "bold"))
)
```

```{r cov_sage_change}
#| fig-cap: "Figure 7. Change in percent cover in sagebrush ecosystems."
#| fig-alt: "Figure 7. Change in percent cover in sagebrush ecosystems."
#| fig-height: 8.5

# Final figure
gridExtra::grid.arrange(
  grobs = cov_sage_figs,
  ncol = 2,
  top = grid::textGrob("Change in Percent Cover in Sagebrush Ecosystmes\n2020-2023 average minus 1990-1993 average",
                       gp = grid::gpar(fontsize = 14, fontface = "bold"))
)
```


# References

Brown, A. (2025). rapr: Interface to Rangeland Analysis Platform (RAP) Vegetation Biomass and Cover Products. R package version 1.0.0. <http://github.com/brownag/rapr>.

Posit team. 2024. RStudio: Integrated Development Environment for R. Posit Software, PBC, Boston, MA. URL <http://www.posit.co/>.

R Core Team. `r R.version$year`. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL <https://www.R-project.org/>.

USDA Agricultural Research Service. 2025. Rangeland Analysis Platform. <https://rangelands.app/>. Accessed 20 May 2025.


# Session Info {#sec-session}

This data-driven report was generated using [Quarto](https://quarto.org/) (version `r quarto::quarto_version()`) and R (version `r paste(R.version$major, R.version$minor, sep = ".")`; R Core Team `r R.version$year`) in RStudio (version 2024.4.2.764; Posit team 2024).
The following output contains information about the environment this script was run in, including the operating system, R version, and packages used.

```{r session, results='hold'}
sessionInfo()
```
